---
title: "GROUP4_MR"
author: "Sandra"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Load packages required for the MR {. tabset}

```{r, packages, include=FALSE, comment=NA}
library(MRInstruments)
library(TwoSampleMR)
```

## Loading in the exposure dataset 

```{r, Exposure data,comment=NA}

# Specify the file path
file_path <- "/home/sandra/Documents/EDA_DATA/Mendelian Randomization/mrc_uganda_hackathon_202306_group4/data/yengo_BMI_656.txt"

#Reading in the exposure data using function read_exposure
bmi_exp_data <- read_exposure_data(
    filename = file_path,
    sep = "\t",
    snp_col = "SNP",
    beta_col = "BETA",
    se_col = "SE",
    effect_allele_col = "Tested_Allele",
    other_allele_col = "Other_Allele",
    eaf_col = "Freq_Tested_Allele_in_HRS",
    pval_col = "P",
    samplesize_col = "N"
)

#How many rows and columns are in the dataset

dim(bmi_exp_data)
head(bmi_exp_data)

```

## Clumping the exposure data

```{r, Clumped data,  comment=NA}
Bmi_exp_data <- clump_data(bmi_exp_data)
#364 variants have been removed due to LD with other variants or absence from the LD reference panel

dim(Bmi_exp_data)

#Renaming the column for Exposure to BMI

Bmi_exp_data$exposure = "BMI"

```



## Loading in the outcome dataset 
```{r, Outcome_data, comment=NA}

file_path = "/home/sandra/Documents/EDA_DATA/Mendelian Randomization/mrc_uganda_hackathon_202306_group4/data/PANUKB_BMI_YENGO_BMI_SNPs.txt"
outcome_data <- read_outcome_data(
    snps = bmi_exp_data$SNP,
    filename = file_path,
    sep = "\t",
    snp_col = "rsid",
    beta_col = "beta_AFR",
    se_col = "se_AFR",
    effect_allele_col = "alt",
    other_allele_col = "ref",
    eaf_col = "af_AFR",
    pval_col = "neglog10_pval_AFR",
)

dim(outcome_data)

outcome_data$outcome = "BMI"

```



## Harmonizing the data {.tabset}

```{r, Harmonized data, comment=NA}
H_data <- harmonise_data(
    exposure_dat = Bmi_exp_data, 
    outcome_dat = outcome_data
)

dim(H_data)
head(H_data)
```


## Checking for palindromic SNPs {.tabset}


```{r, Palidromic SNPs, comment=NA}

palindromic_at <- subset(H_data,effect_allele.outcome%in% "A" & other_allele.outcome %in% "T")
palindromic_ta <- subset(H_data,effect_allele.outcome %in% "T" & other_allele.outcome%in% "A")
palindromic_gc <- subset(H_data,effect_allele.outcome %in% "G" & other_allele.outcome %in% "C")
palindromic_cg <- subset(H_data,effect_allele.outcome %in% "C" & other_allele.outcome %in% "G")
dim(palindromic_at)
dim(palindromic_ta)
dim(palindromic_gc)
dim(palindromic_cg)

```


## Scatter plot for the beta estimate of the exposure and outcome to access for directionality 

```{r, comment=NA}

# Create the scatter plot
plot(H_data$beta.exposure, H_data$beta.outcome, pch = 16, col = "blue", xlab = "Beta_exposure", ylab = "Beta_outcome")

# Add a horizontal line
abline(h = 0, col = "red")

# Add a vertical line
abline(v = 0, col = "green")
```

## Filtering for only SNPs with positive and negative beta estimates to obtain SNPs that are African associated {.tabset}

```{r, comment=NA}
keep <- which(H_data$beta.exposure > 0 & H_data$beta.outcome > 0)
keep <- which(H_data$beta.exposure < 0 & H_data$beta.outcome < 0)
H_data = H_data[keep,]
head(H_data)
dim(H_data)
View(H_data)
```


